Computing properties of independent sets by generic programming tensor networks \thanks{\funding{...}}

### 2021-09-16 paper/paper.tex
fix appendix ABC

\section{Technical guide}\label{sec:technical}
\begin{description}
	\item[OMEinsum] a package for the \texttt{einsum} function,
	\item[OMEinsumContractionOrders] a package for finding the optimal contraction order for the \texttt{einsum} function \\ \href{https://github.com/Happy-Diode/OMEinsumContractionOrders.jl}{https://github.com/Happy-Diode/OMEinsumContractionOrders.jl},
	\item[TropicalGEMM] a package for efficient tropical matrix multiplication (compatible with OMEinsum),
	\item[TropicalNumbers] a package providing tropical number types and tropical algebra, one o the dependency of TropicalGEMM,
	\item[LightGraphs] a package providing graph utilities, like random regular graph generator,
	\item[Polynomials] a package providing polynomial algebra and polynomial fitting,
	\item[Mods and Primes] packages providing finite field algebra and prime number generators.
\end{description}

One can install these packages by opening a Julia REPL, type \colorbox{lightgray}{\texttt{]}} to enter the \texttt{pkg>} mode and type, e.g.
\begin{lstlisting}
pkg> add OMEinsum LightGraphs Mods Primes FFTW Polynomials TropicalNumbers
\end{lstlisting}

It may surprise you that the Julia implementation of algorithms introduced in the paper is so short that except the bounding and sparsity related parts,
all are contained in this appendix. After installing required packages, one can open a Julia REPL and copy the following code into it.

@@

In the above examples, the configuration enumeration is very slow, one should use the optimal MIS size for bounding as described in the main text.
We will not show any example about implementing the backward rule here because it has approximately 100 lines of code.
Please checkout our GitHub repository \href{https://github.com/Happy-Diode/NoteOnTropicalMIS}{https://github.com/Happy-Diode/NoteOnTropicalMIS}.

\section{Why not introducing \texorpdfstring{$\delta$} tensors}\label{app:tensorbad}

@@

Its traditional tensor network representation with $\delta$ tensors is

\centerline{\begin{tikzpicture}
    \def\r{0.08}
    \def\a{0.1}
    \foreach \x in {1,...,5}
        \foreach \y in {1,...,5}
            \filldraw[fill=black] (\x,\y) circle [radius=0.7*\r];
    \foreach \x in {1,...,5}
        \foreach \y in {1,...,4}{
            \filldraw[fill=black] (\x,\y+0.5) circle [radius=\r];
            \filldraw[fill=black] (\y+0.5,\x) circle [radius=\r];
            \draw [black,thick] (\x,\y) -- (\x,\y+1);
            \draw [black,thick] (\y,\x) -- (\y+1,\x);
        }
    \foreach \x in {1,...,4}
        \foreach \y in {1,...,4}{
            \filldraw[fill=black] (\x+0.3,\y+0.3) circle [radius=\r];
            \filldraw[fill=black] (\y+0.3,\x+0.7) circle [radius=\r];
            \draw [black,thick] (\x,\y) -- (\x+1,\y+1);
            \draw [black,thick] (\y+1,\x) -- (\y,\x+1);
        }
    \tikzset{decoration={snake,amplitude=.4mm,segment length=2mm,
                    post length=0mm,pre length=0mm}}
    \draw [decorate] (2.7, 0.5) -- (2.7, 5.5);
\end{tikzpicture}}
where a small circle on an edge is a diagonal tensor. Its rank is $8$ in the bulk. If we contract this tensor network in a naive column-wise order, the maximum intermediate tensor is approximately $3L$, giving a space complexity $\approx 2^{3L}$.
If we treat it as the following generalized tensor network

\centerline{\begin{tikzpicture}
    \def\r{0.08}
    \def\a{0.07}
    \def\L{0.6}
    \def\l{0.1}
    \def\sql{0.24}
    \pgfmathsetseed{2}
    \foreach[evaluate={\cr=0.1+0.5*Mod(\x,2)}] \x in {1,...,5}
        \foreach[evaluate={\cg=0.1+0.3*Mod(\y,2); \cy=0.5-0.5*Mod(\y,2)}] \y in {1,...,5}{
            \edef\R{\pdfuniformdeviate 255}
            \edef\G{\pdfuniformdeviate 255}
            \edef\B{\pdfuniformdeviate 255}
            \xdefinecolor{MyColor}{RGB}{\R,\G,\B}
            \ifnum \x < 5
                \draw [thick, MyColor, opacity=1.0, line cap=round] (\x,\y) -- (\x+0.5,\y);
                \ifnum \y < 5
                \draw [thick, MyColor, opacity=1.0, line cap=round] (\x,\y) -- (\x+0.3,\y+0.3);
                \fi
                \ifnum \y > 1
                \draw [thick, MyColor, opacity=1.0, line cap=round] (\x,\y) -- (\x+0.3,\y-0.3);
                \fi
            \fi
            \ifnum \x > 1
                \draw [thick, MyColor, opacity=1.0, line cap=round] (\x,\y) -- (\x-0.5,\y);
                \ifnum \y < 5
                \draw [thick, MyColor, opacity=1.0, line cap=round] (\x,\y) -- (\x-0.7,\y+0.7);
                \fi
                \ifnum \y > 1
                \draw [thick, MyColor, opacity=1.0, line cap=round] (\x,\y) -- (\x-0.7,\y-0.7);
                \fi
            \fi
            \ifnum \y < 5
                \draw [thick, MyColor, opacity=1.0, line cap=round] (\x,\y) -- (\x,\y+0.5);
            \fi
            \ifnum \y > 1
                \draw [thick, MyColor, opacity=1.0, line cap=round] (\x,\y) -- (\x,\y-0.5);
            \fi
        }
    \foreach \x in {1,...,5}
        \foreach \y in {1,...,5}{
            \filldraw[fill=black] (\x,\y) circle [radius=0.7*\r];
        }
    \foreach \x in {1,...,5}
        \foreach \y in {1,...,5}{
            \ifnum \y < 5
                \filldraw[fill=black] (\x,\y+0.5) circle [radius=\r];
                \filldraw[fill=black] (\y+0.5,\x) circle [radius=\r];
            \fi
        }
    \foreach \x in {1,...,4}
        \foreach \y in {1,...,4}{
            \filldraw[fill=black] (\x+0.3,\y+0.3) circle [radius=\r];
            \filldraw[fill=black] (\y+0.3,\x+0.7) circle [radius=\r];
        }
    \tikzset{decoration={snake,amplitude=.4mm,segment length=2mm,
                    post length=0mm,pre length=0mm}}
    \draw [decorate] (2.7, 0.5) -- (2.7, 5.5);
\end{tikzpicture}}
where we use different colors to distinguish different hyperedges.
Now, the vertex tensor is always rank $1$.
With the same naive contraction order, we can see the maximum intermediate tensor is approximately of size $2^L$ by counting the colors.

\section{Generalizing to other graph problems}\label{app:otherproblems}
There are some other graph problems that can be encoded in a tensor network.
To understand its representation power, it is a good starting point to connect it with dynamic programming because
a tensor network can be viewed as a special type of dynamic programming where its update rule can be characterized by a linear operation.
Courcelle’s theorem~\cite{Courcelle1990,Barr2020} states that a problem quantified by monadic second order logic (MSO) on a graph with bounded treewidth $k$ can be solved in linear time with respect to the graph size.
Dynamic programming is a traditional approach to attack a MSO problem, it can solve the maximum independent set problem in $O(2^k)n$, which is similar to the tensor network approach.
We mentioned in the main text that tensor network has nice analytic property make it easier for generic programming.
The cost is, the tensor network is less expressive than dynamic programming,
%\iffalse
%The cost is, it is less expressive than MSO because the tensor network described by \Eq{eq:tensor} can be expressed in MSO as
%\begin{align}
%    \begin{split}
%    \exists_X\forall_{u}&(\\
%    &\quad\forall_{v} 
%    \neg {\rm adj}(u, v) \lor
%    ({\rm adj}(u, v) \land ( \hspace{5em}\text{$\triangleright$ restrictions on edges}\\
%    &\quad\quad(u \not\in X \land v \not\in X \land B_{00}) \lor\\
%    &\quad\quad(u \not\in X \land v \in X \land B_{01}) \lor\\
%    &\quad\quad(u \in X \land v \not\in X \land B_{10}) \lor\\
%    &\quad\quad(u \in X \land v \in X \land B_{11})\\
%    &\quad)\\
%    &))\land\\
%    &(\hspace{16.5em}\text{$\triangleright$ restrictions on vertices}\\
%    &\quad(u \not\in X \land W_{0}) \lor\\
%    &\quad(u \in X \land W_{1})\\
%    &),
%    \end{split}
%\end{align}
%while not all monadic second order logic can be represented as a tensor network contraction,
%for example, it is hard to construct a tensor network to decide whether a graph is connected or not.
%At the cost of losing expressiveness, we can encode the properties of the graph into the tensor elements.
%\fi
However, that are still some other problems that can be expressed in the framework of generic tensor network.
\subsection{Matching problem}
A matching polynomial of a graph $G$ is defined as
\begin{equation}
    M(G, x) = \sum\limits_{k=1}^{|V|/2} c_k x^k,
\end{equation}
where $k$ is the number of matches, and coefficients $c_k$ are the corresponding counting.

We define a tensor of rank $d(v) = |N(v)|$ on vertex $v$ such that,
\begin{equation}
    W_{v\rightarrow n_1, v\rightarrow n_2, \ldots, v\rightarrow n_{d(v)}} = \begin{cases}
        1, & \sum_{i=1}^{d(v)} v\rightarrow n_i \leq 1,\\
        0, & \text{otherwise},
    \end{cases}
\end{equation}
and a tensor of rank $1$ on the bond
\begin{equation}
    B_{v\rightarrow w} = \begin{cases}
    1, & v \rightarrow w = 0 \\
    x, & v \rightarrow w = 1.
\end{cases}
\end{equation}
Here, we use bond index $v \rightarrow w$ to label tensors.

\subsection{k-Coloring}
Let us use 3-coloring on the vertex as an example. We can define a vertex tensor as
\begin{equation}
    W = \left(\begin{matrix}
        r_v\\
        g_v\\
        b_v
    \end{matrix}\right),
\end{equation}
and an edge tensor as
\begin{equation}
    B = \left(\begin{matrix}
        0 & 1 & 1\\
        1 & 0 & 1\\
        1 & 1 & 0
    \end{matrix}\right).
\end{equation}
The number of possible coloring can be obtained by contracting this tensor network by setting vertex tensor elements $r_v, g_v$ and $b_v$ to $1$.
By designing generic types as tensor elements, one should be able to get all possible colorings.
It is straight forward to define the k-coloring problem on edges hence we will not discuss the detailed construction here.

\subsection{Max cut problem}
Max cut problem is also known as the boolean spinglass problem.
Its tensor network representation does not contain vertex tensors.
The edge tensor can be defined as
\begin{equation}
    B(x_{ij}) = \left(\begin{matrix}
        1 & x_{ij}\\
        x_{ij} & 1
    \end{matrix}\right).
\end{equation}
We can also define a polynomial about edges variables by setting $x_{ij} = x$,
its $k$th coefficient is two times the number of configurations of cut size $k$.

\subsection{Set packing}
Set packing the the hypergraph version of the maximum independent set problem, where each set corresponds to a vertex, each element corresponds to a hyperedge.
To solve the set packing problem, we just remove the rank 2 restriction of an edge tensor
\begin{equation}
    B_{v,w,\ldots, z} = \begin{cases}
        1, & v+w+\ldots+z\leq 1,\\
        0, & otherwise
    \end{cases}
\end{equation}

>>>>

\section{Technical guide}\label{sec:technical}
\begin{description}
	\item[\href{https://github.com/under-Peter/OMEinsum.jl}{OMEinsum} and \href{https://github.com/Happy-Diode/OMEinsumContractionOrders.jl}{OMEinsumContractionOrders}] are packages providing the support for Einstein's (or tensor network) notation and contraction order optimizations.
    \texttt{OMEinsumContractionOrders} implements state of art algorithms for finding the optimal contraction order for the an tensor network, including the KaHypar+Greedy~\cite{Gray2021, Pan2021} and local transformation based approaches~\cite{Kalachev2021},
	\item[\href{https://github.com/TensorBFS/TropicalNumbers.jl}{TropicalNumbers} and \href{https://github.com/TensorBFS/TropicalGEMM.jl}{TropicalGEMM}] are packages providing tropical number and efficient tropical matrix multiplication,
	\item[\href{https://github.com/JuliaGraphs/LightGraphs.jl}{LightGraphs}] is a package providing graph utilities, like random regular graph generator,
	\item[\href{https://github.com/JuliaMath/Polynomials.jl}{Polynomials}] is a package providing polynomial algebra and polynomial fitting,
	\item[\href{https://github.com/scheinerman/Mods.jl}{Mods} and \href{https://github.com/JuliaMath/Primes.jl}{Primes}] are packages providing finite field algebra and prime number generations.
\end{description}

One can install these packages by opening a Julia REPL, type \colorbox{lightgray}{\texttt{]}} to enter the \texttt{pkg>} mode and type, e.g.
\begin{lstlisting}
pkg> add OMEinsum LightGraphs Mods Primes Polynomials TropicalNumbers OMEinsumContractionOrders
\end{lstlisting}

It may surprise you that the Julia implementation of algorithms introduced in the paper is so short that except the bounding algorithm,
all are contained in this appendix. After installing required packages, one can open a Julia REPL and copy the following code into it.

@@

Please checkout our GitHub repository for the full featured version:

\href{https://github.com/Happy-Diode/GraphTensorNetworks.jl}{https://github.com/Happy-Diode/GraphTensorNetworks.jl}

\section{Why not using the standard tensor network notations}\label{app:tensorbad}
As we have mentioned in the main text,
a standard tensor network notation inevitably introduces $\delta$ tensors to encode some graph problems.
A $\delta$ tensor of rank $d$ is defined as
\begin{equation}
    \delta_{i_1, i_2,\ldots,i_d} = \begin{cases}
        1, & i_1=i_2=\ldots =i_d,\\
        0, & \text{otherwise}.
    \end{cases}
\end{equation}

@@

Its standard tensor network representation includes a $\delta$ tensors on each vertex, so its graphical representation is as bellow.

\centerline{\begin{tikzpicture}
    \def\r{0.08}
    \def\a{0.1}
    \foreach \x in {1,...,5}
        \foreach \y in {1,...,5}
            \filldraw[fill=black] (\x,\y) circle [radius=0.7*\r];
    \foreach \x in {1,...,5}
        \foreach \y in {1,...,4}{
            \filldraw[fill=black] (\x,\y+0.5) circle [radius=\r];
            \filldraw[fill=black] (\y+0.5,\x) circle [radius=\r];
            \draw [black,thick] (\x,\y) -- (\x,\y+1);
            \draw [black,thick] (\y,\x) -- (\y+1,\x);
        }
    \foreach \x in {1,...,4}
        \foreach \y in {1,...,4}{
            \filldraw[fill=black] (\x+0.3,\y+0.3) circle [radius=\r];
            \filldraw[fill=black] (\y+0.3,\x+0.7) circle [radius=\r];
            \draw [black,thick] (\x,\y) -- (\x+1,\y+1);
            \draw [black,thick] (\y+1,\x) -- (\y,\x+1);
        }
    \tikzset{decoration={snake,amplitude=.4mm,segment length=2mm,
                    post length=0mm,pre length=0mm}}
    \draw [decorate] (2.7, 0.5) -- (2.7, 5.5);
\end{tikzpicture}}
Here, we use a small circle on a vertex to represent a diagonal $\delta$ tensor. Its rank is $8$ in the bulk. If we contract this tensor network in a naive column-wise order, the maximum intermediate tensor has a rank $\sim3L$, requiring a storage of size $\approx 2^{3L}$.
If we treat it as the following generalized tensor network

\centerline{\begin{tikzpicture}
    \def\r{0.08}
    \def\a{0.07}
    \def\L{0.6}
    \def\l{0.1}
    \def\sql{0.24}
    \pgfmathsetseed{2}
    \foreach[evaluate={\cr=0.1+0.5*Mod(\x,2)}] \x in {1,...,5}
        \foreach[evaluate={\cg=0.1+0.3*Mod(\y,2); \cy=0.5-0.5*Mod(\y,2)}] \y in {1,...,5}{
            \edef\R{\pdfuniformdeviate 255}
            \edef\G{\pdfuniformdeviate 255}
            \edef\B{\pdfuniformdeviate 255}
            \xdefinecolor{MyColor}{RGB}{\R,\G,\B}
            \ifnum \x < 5
                \draw [thick, MyColor, opacity=1.0, line cap=round] (\x,\y) -- (\x+0.5,\y);
                \ifnum \y < 5
                \draw [thick, MyColor, opacity=1.0, line cap=round] (\x,\y) -- (\x+0.3,\y+0.3);
                \fi
                \ifnum \y > 1
                \draw [thick, MyColor, opacity=1.0, line cap=round] (\x,\y) -- (\x+0.3,\y-0.3);
                \fi
            \fi
            \ifnum \x > 1
                \draw [thick, MyColor, opacity=1.0, line cap=round] (\x,\y) -- (\x-0.5,\y);
                \ifnum \y < 5
                \draw [thick, MyColor, opacity=1.0, line cap=round] (\x,\y) -- (\x-0.7,\y+0.7);
                \fi
                \ifnum \y > 1
                \draw [thick, MyColor, opacity=1.0, line cap=round] (\x,\y) -- (\x-0.7,\y-0.7);
                \fi
            \fi
            \ifnum \y < 5
                \draw [thick, MyColor, opacity=1.0, line cap=round] (\x,\y) -- (\x,\y+0.5);
            \fi
            \ifnum \y > 1
                \draw [thick, MyColor, opacity=1.0, line cap=round] (\x,\y) -- (\x,\y-0.5);
            \fi
        }
    \foreach \x in {1,...,5}
        \foreach \y in {1,...,5}{
            \filldraw[fill=black] (\x,\y) circle [radius=0.7*\r];
        }
    \foreach \x in {1,...,5}
        \foreach \y in {1,...,5}{
            \ifnum \y < 5
                \filldraw[fill=black] (\x,\y+0.5) circle [radius=\r];
                \filldraw[fill=black] (\y+0.5,\x) circle [radius=\r];
            \fi
        }
    \foreach \x in {1,...,4}
        \foreach \y in {1,...,4}{
            \filldraw[fill=black] (\x+0.3,\y+0.3) circle [radius=\r];
            \filldraw[fill=black] (\y+0.3,\x+0.7) circle [radius=\r];
        }
    \tikzset{decoration={snake,amplitude=.4mm,segment length=2mm,
                    post length=0mm,pre length=0mm}}
    \draw [decorate] (2.7, 0.5) -- (2.7, 5.5);
\end{tikzpicture}}
Here, we use different colors to distinguish different hyperedges.
The vertex tensor is always rank $1$ in this representation.
Again, if we contract this tensor network in the columnwise contraction order, we can see the maximum intermediate tensor has rank $\sim L$ by counting the colors.

\section{Generalizing to other graph problems}\label{app:otherproblems}
There are some other graph problems that can be encoded to a tensor network.
To understand its representation power, it is a good starting point to connect it with dynamic programming because
a tensor network can be viewed as a special type of dynamic programming where its update rule can be characterized by a linear operation.
Dynamic programming can solve monadic second order logic (MSO) definable problem on a graph with small tree width efficiently~\cite{Courcelle1990,Barr2020}.
It can solve the maximum independent set problem in $O(2^k)n$, which is similar to the tensor network approach.
The graph problem can be expressed in the tensor network language is clearly much less than that can be expressed by dynamic programming.
For example, it is hard to determine whether a collection of vertices forms a path in a graph using the tensor network language, while this problem is MSO definable.
As a gain, a tensor network has a nicer analytic property, which makes contraction order optimization, generic programming and utilizing BLAS libaries very convenient.
%\iffalse
%The cost is, it is less expressive than MSO because the tensor network described by \Eq{eq:tensor} can be expressed in MSO as
%\begin{align}
%    \begin{split}
%    \exists_X\forall_{u}&(\\
%    &\quad\forall_{v} 
%    \neg {\rm adj}(u, v) \lor
%    ({\rm adj}(u, v) \land ( \hspace{5em}\text{$\triangleright$ restrictions on edges}\\
%    &\quad\quad(u \not\in X \land v \not\in X \land B_{00}) \lor\\
%    &\quad\quad(u \not\in X \land v \in X \land B_{01}) \lor\\
%    &\quad\quad(u \in X \land v \not\in X \land B_{10}) \lor\\
%    &\quad\quad(u \in X \land v \in X \land B_{11})\\
%    &\quad)\\
%    &))\land\\
%    &(\hspace{16.5em}\text{$\triangleright$ restrictions on vertices}\\
%    &\quad(u \not\in X \land W_{0}) \lor\\
%    &\quad(u \in X \land W_{1})\\
%    &),
%    \end{split}
%\end{align}
%while not all monadic second order logic can be represented as a tensor network contraction,
%for example, it is hard to construct a tensor network to decide whether a graph is connected or not.
%At the cost of losing expressiveness, we can encode the properties of the graph into the tensor elements.
%\fi
In the following, we introduce some other problems that can be expressed by a tensor network.
\subsection{Matching problem}
A matching polynomial of a graph $G$ is defined as
\begin{equation}
    M(G, x) = \sum\limits_{k=1}^{|V|/2} c_k x^k,
\end{equation}
where $k$ is the number of matches, and coefficients $c_k$ are the corresponding counting.
We define a tensor of rank $d(v) = |N(v)|$ on vertex $v$ such that,
\begin{equation}
    W_{\langle v, n_1\rangle, \langle v, n_2 \rangle, \ldots, \langle v, n_{d(v)}\rangle} = \begin{cases}
        1, & \sum_{i=1}^{d(v)} \langle v, n_i \rangle \leq 1,\\
        0, & \text{otherwise},
    \end{cases}
\end{equation}
and a tensor of rank $1$ on the bond
\begin{equation}
    B_{\langle v, w\rangle} = \begin{cases}
    1, & \langle v, w \rangle = 0 \\
    x, & \langle v, w \rangle = 1.
\end{cases}
\end{equation}
Here, we use bond index $\langle v, w \rangle \in \{0, 1\}$ ($\langle v, w \rangle$ and $\langle w,v\rangle$ are equivalent) to label tensors,
where $1$ means two vertices of an edge are matched, $0$ means otherwise.

\subsection{k-Coloring}
Let us use 3-coloring on the vertex as an example. We can define a vertex tensor as
\begin{equation}
    W = \left(\begin{matrix}
        r_v\\
        g_v\\
        b_v
    \end{matrix}\right),
\end{equation}
and an edge tensor as
\begin{equation}
    B = \left(\begin{matrix}
        0 & 1 & 1\\
        1 & 0 & 1\\
        1 & 1 & 0
    \end{matrix}\right).
\end{equation}
The number of possible coloring can be obtained by contracting this tensor network by setting vertex tensor elements $r_v, g_v$ and $b_v$ to $1$.
By designing generic types as tensor elements, one can get all possible colorings.
Similarly, one can define the k-coloring problem on edges too.

\subsection{Max cut problem}
Max cut problem is also known as the boolean spinglass problem.
Its tensor network representation does not contain vertex tensors.
The edge tensor can be defined as
\begin{equation}
    B(x_{\langle i, j\rangle}) = \left(\begin{matrix}
        1 & x_{\langle i, j\rangle}\\
        x_{\langle i, j\rangle} & 1
    \end{matrix}\right).
\end{equation}
We can also define a polynomial about edges variables by setting $x_{\langle i, j\rangle} = x$,
where its $k$th coefficient is two times the number of configurations of cut size $k$.

\subsection{Set packing}
Set packing is the hypergraph version of the maximum independent set problem, where each set corresponds to a vertex, each element corresponds to a hyperedge.
To solve the set packing problem, we just remove the rank 2 restriction of an edge tensor
\begin{equation}
    B_{v,w,\ldots, z} = \begin{cases}
        1, & v+w+\ldots+z\leq 1,\\
        0, & otherwise
    \end{cases}
\end{equation}
